---
title: "BIOS 731 Homework 4: Bayesian Inference and Cluster Computing"
format: 
  pdf:
    number-depth: 4
    mainfont: Times New Roman
    include-in-header:
      - text: |
          \usepackage{makecell}
          \usepackage{longtable}
          \usepackage{mathrsfs}
    df-print: kable
    knitr:
      opts_chunk:
        echo: true,
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        fig.path: "../results/"
        R.options:
          scipen: 999
          knitr.kable.NA: ""
editor: source
---

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc()

# Load packages
# if(!require("here")) install.packages("here")
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  gtools, DescTools, MASS,
  tidyverse, magrittr, janitor,
  knitr, kableExtra, broom, here, tictoc
)
```

```{r}
#| label: source functions

source(here("source", "gibbs_sampler_fn.R"))
source(here("source", "variational_inference_fn.R"))
```


# Problem 0: Reproducibility

The GitHub repository containing this work can be found at

<https://github.com/ryanjtaylor97/bios731_hw4_taylor>.

# Problem 1: Gibbs sampler

We have the data-generating model:

$$
\begin{aligned}
\mu & \sim N(0, \sigma^2) \\
c_i & \sim Categorical(1/K, \ldots, 1/K) \\
y_i | c_i, \mu & \sim N(c_i^\top \mu, 1).
\end{aligned}
$$

Then first, for the full conditional distribution $p(c_i | y_i, \mu)$, note that $c_i$ can take on one of $K$ different values: $(1,0,\ldots,0); \ (0, 1, 0, \ldots, 0); \ldots; (0, \ldots, 0, 1).$ Each of these possibilities has probability $p(c_{ik} = 1) = {1 \choose 1! * (0!)^{K-1}} * (\frac{1}{K})^1 * (\frac{1}{K})^{K-1} = \frac{1}{K}$. In addition, when $c_{ik} = 1, c_i^\top \mu = \mu_k$. Therefore, we can write

$$
\begin{aligned}
p(c_{ik}=1 | y_i, \mu) & \propto p(c_{ik} = 1) * p(y_i | c_{ik} = 1, \mu) \\
& \propto exp\Big( -\frac{1}{2} (y_i - \mu_k)^2 \Big) * \frac{1}{K} \\
\implies [c_i | y_i, \mu] & \sim Categorical \Bigg( \frac{\phi(y_i; \mu_k, 1)}{\sum_{k = 1}^K \phi(y_i; \mu_k, 1) } \Bigg),
\end{aligned}
$$

where $\phi(\cdot; \mu, \sigma^2)$ is the pdf of a univariate Normal distribution with mean $\mu$ and variance $\sigma^2$.

Second, since we assume all $y_i$'s are independent of each other, we can derive an update all $\mu_k$ parameters at the same time using the entire data vector $\boldsymbol{y}$ and the entire $n \times K$ matrix of group assignments $\boldsymbol{c}$.

$$
\begin{aligned}
p(\boldsymbol{\mu} | \boldsymbol{y}, \boldsymbol{c}) & \propto p(\boldsymbol{y} | \boldsymbol{c}, \boldsymbol{\mu}) * p(\boldsymbol{\mu}) \\
& \propto exp \Big( -\frac{1}{2} (\boldsymbol{y} - \boldsymbol{c}^\top \boldsymbol{\mu})^\top (\boldsymbol{y} - \boldsymbol{c}^\top \boldsymbol{\mu}) \Big) * exp \Big( -\frac{1}{2} \boldsymbol{\mu}^\top \boldsymbol{\mu} \Big) \\
&\propto exp \Big\{ -\frac{1}{2} \Big[ \boldsymbol{\mu}^\top \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) \boldsymbol{\mu} - 2 \boldsymbol{\mu}^\top \boldsymbol{c}^\top \boldsymbol{y} \Big] \Big\} \\
& \propto exp \Big\{ -\frac{1}{2} \Big[ \boldsymbol{\mu} - \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) ^{-1} \Big) \boldsymbol{c}^\top \boldsymbol{y} \Big]^\top \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) \Big[ \boldsymbol{\mu} - \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) ^{-1} \Big) \boldsymbol{c}^\top \boldsymbol{y} \Big] \Big\} \\
\implies [\boldsymbol{\mu} | \boldsymbol{y}, \boldsymbol{c}] & \sim N \Bigg( \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) ^{-1} \Big) \boldsymbol{c}^\top \boldsymbol{y}, \Big( \boldsymbol{c}^\top \boldsymbol{c} + \frac{1}{\sigma^2} \mathbf{I}_K \Big) ^{-1} \Big) \Bigg).
\end{aligned}
$$

Since $\boldsymbol{c}$ is a matrix of indicator vectors, $\boldsymbol{c}^\top \boldsymbol{c}$ is a diagonal matrix, and therefore we can rewrite this as $K$ univariate normal distributions so we do not need to invert a matrix:

$$
\boldsymbol{c}^\top \boldsymbol{c} = \begin{bmatrix}
n_1 & 0 & 0 \\
0 & n_2 & 0 \\
0 & 0 & n_3
\end{bmatrix},
$$

where $n_k = \sum_{i=1}^n c_{ik}$. Therefore, we can write $[\mu_k | \boldsymbol{y}, \boldsymbol{c}] \sim N(\frac{\sum_{i}^n c_{ik}y_i}{n_k + 1/\sigma^2}, \frac{1}{n_k + 1/\sigma^2})$.

These full conditionals are implemented in the Gibbs sampler in the file `source\gibbs_sampler_fn.R`.

# Problem 2: Variational inference

We need to estimate the "latent" parameters $\boldsymbol{\mu}, \boldsymbol{c}$. In order to estimate these, we'll also define the variational parameters $\boldsymbol{m} = \mathbb{E}_q(\boldsymbol{\mu}), \ \boldsymbol{s}^2 = Var_q(\boldsymbol{\mu}), \text{ and } \boldsymbol{\psi} = \mathbb{E}_q(\boldsymbol{c})$, so that $\psi_{ik} = P(c_{ik} = 1)$. Then by a mean-field approximation, we can write 

$$
\begin{aligned}
ELBO(q(\boldsymbol{\mu}, \boldsymbol{c})) = {} & E_q[log \ p(y, \mu, c)] - E_q[log \ q(\mu, c)] \\
= {} & E_q[log \ p(y | \mu, c) + log \ p(\mu) + log \ p(c)] - E_q[log \ q(\mu) + log \ q(c)] \\
= {} & \sum_{i=1}^n \{ E_q [ log \ p(y_i|c_i, \mu)] + E_q[log \ p(c_i)] \} + \sum_{k=1}^K E_q[log \ p(\mu_k)] - \\
& \sum_{i=1}^n E_q[log \ q(c_i)] - \sum_{k=1}^K E_q[log \ q(\mu_k)].
\end{aligned}
$$

Then by the formula we proved in class, the CAVI update for $c_i$ is 

$$
\begin{aligned}
q^*(c_i) & \propto exp \Big\{ log \ p(c_i) + \mathbb{E}_{q(\mu)}[log \ p(x_i | c_i, \mu)] \Big\} \\
& \propto exp \Big\{ -log(K) + \mathbb{E}_{q(\mu)} \Big[\sum_{k=1}^K c_{ik} *(-\frac{1}{2} (x_i - \mu_k)^2 \Big] \Big\} \\
& \propto exp \Big\{ \sum_{k=1}^K c_{ik} * \mathbb{E}_{q(\mu)} (-\frac{1}{2} (x_i - \mu_k)^2) \Big\} \\
& \propto exp \Big\{ \sum_{k=1}^K c_{ik} * \Big[ E(\mu_k) x_i - \frac{1}{2} E(\mu_k^2) \Big] \Big\} \\
&= exp \Big\{ \sum_{k=1}^K c_{ik} * \Big[ m_k x_i - \frac{1}{2} (s^2_k + m_k^2) \Big] \Big\} \\
\implies \psi_{ik}^* & \propto exp \Big[ m_k x_i - \frac{1}{2} (s^2_k + m_k^2) \Big]
\end{aligned}
$$

Likewise, the CAVI update for $\mu_k$ is 

$$
\begin{aligned}
q^*(\mu_k) & \propto exp \Big\{ log \ p(\mu_k) + \mathbb{E}_{q(c)}[log \ p(x_i | c_i, \mu)] \Big\} \\
& \propto exp \Big\{ -\frac{\mu_k^2}{2 \sigma^2} + \sum_{i=1}^n \mathbb{E}_{q(c)} [ c_{ik} * log \ p(x_i | \mu_k) ] \Big\} \\
& \propto exp \Big\{ -\frac{\mu_k^2}{2 \sigma^2} + \sum_{i=1}^n \Big[ \psi_{ik} * \Big( -\frac{1}{2} * (x_i - \mu_k)^2 \Big) \Big] \Big\} \\
& \propto exp \Big\{ \mu_k^2 \Big[ \frac{-1}{2 \sigma^2} + \sum_{i=1}^n \frac{-\psi_{ik}}{2} \Big] + \mu_k \Big[ \sum_{i=1}^n \psi_{ik} x_i \Big] \Big\} \\
& \propto exp \Big\{ \frac{-1}{2} \Big( \frac{1}{\sigma^2} + \sum_{i=1}^n \psi_{ik} \Big) * \Big[ \mu_k^2 + 2 \mu_k \Big( \frac{1}{\sigma^2} + \sum_{i=1}^n \psi_{ik} \Big)^{-1} \Big( \sum_{i=1}^n \psi_{ik} x_i \Big) + \Big( \frac{1}{\sigma^2} + \sum_{i=1}^n \psi_{ik} \Big)^{-2} \Big( \sum_{i=1}^n \psi_{ik} x_i \Big)^2 \Big] \Big\} \\ \\
\implies q^*(\mu_k) & = N \Bigg( \frac{\sum_{i=1}^n \psi_{ik} x_i}{\frac{1}{\sigma^2} + \sum_{i=1}^n \psi_{ik}}, \frac{1}{\frac{1}{\sigma^2} + \sum_{i=1}^n \psi_{ik}} \Bigg) \\ \\
&= N(m_k, s^2_k).
\end{aligned}
$$

Finally, we can rewrite the ELBO using these updates:

$$
\begin{aligned}
ELBO(q) = {} & \sum_{i=1}^n \{ E_q [ log \ p(y_i|c_i, \mu)] + E_q[log \ p(c_i)] \} + \sum_{k=1}^K E_q[log \ p(\mu_k)] - \\
& \sum_{i=1}^n E_q[log \ q(c_i)] - \sum_{k=1}^K E_q[log \ q(\mu_k)] \\
= {} & \sum_{i=1}^n \{ E_{q(c)}[E_{q(\mu)}[-\frac{1}{2} (y_i - \mu_k)^2] | c] \ - log \ K \} + \sum_{k=1}^K E_q \Big[ -\frac{\mu_k}{2 \sigma^2} \Big] - \\
& \sum_{i=1}^n E_q \Big[ log(\prod_{k=1}^K \psi_{ik}^{c_{ik}}) \Big] - \sum_{k=1}^K E_q \Big[ -\frac{1}{2} log(2 \pi s^2_k) - \frac{1}{2 s^2_k} (\mu_k - m_k)^2 \Big] + \text{constant} \\
={} & \sum_{i=1}^n \sum_{k=1}^K -\frac{\psi_{ik}}{2} [(y_i - m_k)^2 + s^2_k] - \sum_{k=1}^K \frac{1}{2 \sigma^2} (m^2_k + s^2_k) - \sum_{i=1}^n \sum_{k=1}^K \psi_{ik} log(\psi_{ik}) + \sum_{k=1}^K \frac{1}{2} log(2 \pi s^2_k) + \text{const.}
\end{aligned}
$$

We use this definition of the ELBO to monitor convergence in the CAVI algorithm implemented in the file `source\variational_inference_fn.R`.

# Problem 3: Fit to data

```{r}
#| label: load data

data("faithful")

scale_factor <- sd(faithful$waiting)

y_vec <- faithful$waiting / scale_factor

K_expected <- 2
```

```{r}
#| label: run gibbs functions

set.seed(978)

# Run 1 Gibbs chain on Old Faithful data
of_gibbs_100 <- gibbs_k_gauss(
  y = y_vec,
  K = K_expected,
  mu0 = 70 / scale_factor,
  sigma2 = 100
)

of_gibbs_100_summ <- summ_gibbs_mix(of_gibbs_100)

set.seed(617)

# Run second Gibbs chain with lower proposed variance
of_gibbs_10 <- gibbs_k_gauss(
  y = y_vec,
  K = K_expected,
  mu0 = 70 / scale_factor,
  sigma2 = 10
)

of_gibbs_10_summ <- summ_gibbs_mix(of_gibbs_10)

set.seed(508)

# Run third Gibbs chain with even lower proposed variance
of_gibbs_1 <- gibbs_k_gauss(
  y = y_vec,
  K = K_expected,
  mu0 = 70 / scale_factor,
  sigma2 = 1
)

of_gibbs_1_summ <- summ_gibbs_mix(of_gibbs_1)
```

```{r}
#| label: run cavi function

# Run CAVI function on Old Faithful data
of_cavi <- cavi_k_gauss(
  y = y_vec,
  K = K_expected + 4,
  mu0 = c(50, 80) / scale_factor,
  s2_0 = c(10, 10) / scale_factor,
  sigma2 = 100,
  tol = 1e-6
)
```

